Endogenous Macrodynamics in Algorithmic Recourse

### 2022-08-11 dev/paper/paper.tex
some rejigging of sections

\begin{document}
\maketitle
\begin{abstract}
Existing work on Counterfactual Explanations (CE) and Algorithmic
Recourse (AR) has largely been limited to the static setting: given some
classifier we are interested in finding close, actionable, realistic,
sparse, diverse and ideally causally founded counterfactuals. The
ability of CE to handle dynamics like data and model drift remains a
largely unexplored research challenge at this point. Only one recent
work considers the implications of exogenous domain and model shifts.
This project instead focuses on endogenous dynamics, that is shifts that
occur when AR is actually implemented by a proportion of individuals.
Early findings suggest that the involved shifts may be large with
important implications on the validity of AR and the overall
characteristics of the sample population.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, enhanced, sharp corners, breakable, frame hidden, interior hidden]}{\end{tcolorbox}}\fi

@@

\hypertarget{sec-related}{%
\section{Related Work}\label{sec-related}}

@@

\hypertarget{sec-method}{%
\section{Methodology}\label{sec-method}}

In the following we first set out a generalized framework for
gradient-based counterfactual search in Section~\ref{sec-method-general}
to introduce the various counterfactual generators we have chosen to use
in our experiments. We then describe the experimental setup in
Section~\ref{sec-method-experiment} and introduce several evaluation
metrics used to benchmark the different generators.

\hypertarget{sec-method-general}{%
\subsection{A Generalized Framework for Gradient-Based Counterfactual
Search}\label{sec-method-general}}

In this work we have chosen to focus on a number of gradient-based
counterfactual generators to investigate the endogenous dynamics we
introduced in Section~\ref{sec-intro}. Gradient-based counterfactual
search is well-suited for differentiable black-box models like deep
neural networks.

\hypertarget{from-individual-recourse}{%
\subsubsection{From individual recourse
\ldots{}}\label{from-individual-recourse}}

We can restate Equation~\ref{eq-solution} in a more general form that
encompasses most gradient-based approaches to counterfactual search:

@@

Here \(\mathbf{s}^\prime=\left\{s_k^\prime\right\}_K\) is the stacked
\(K\)-dimensional array of counterfactual states and
\(f: \mathcal{S} \mapsto \mathcal{X}\) maps from the counterfactual
state space to the feature space. In the case of the baseline
counterfactual generator
\protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}} \(f\) is just
the idendity function and the number of counterfactuals \(K\) is equal
to one. This generator, which we shall refer to as \textbf{Wachter} in
the following, shall serve as the baseline against which all other
gradient-based methodologies will be compared. In particular, we include
include the following generator in our benchmarking exercises: REVISE
\protect\hyperlink{ref-joshi2019towards}{{[}8{]}}, CLUE
\protect\hyperlink{ref-antoran2020getting}{{[}10{]}}, DICE
\protect\hyperlink{ref-mothilal2020explaining}{{[}9{]}} and a greedy
approach that relies on probabilistic models
\protect\hyperlink{ref-schut2021generating}{{[}7{]}}.

\hypertarget{towards-collective-recourse}{%
\subsubsection{\ldots{} towards collective
recourse}\label{towards-collective-recourse}}

@@

\hypertarget{a-note-on-convergence}{%
\subsubsection{A note on convergence}\label{a-note-on-convergence}}

@@

\hypertarget{sec-method-experiment}{%
\subsection{Experimental Setup}\label{sec-method-experiment}}

@@

\hypertarget{sec-method-data}{%
\subsection{Data}\label{sec-method-data}}

@@

Additionally, we use two real-world datasets from the Finance domain.
Firstly, we use the Give Me Some Credit dataset which was open-sourced
on Kaggle for the task to predict whether a borrower is likely to
experience financial difficulties in the next two years
\protect\hyperlink{ref-gmsc_data}{{[}24{]}}. Originally consisting of
250,000 instances with 11 numerical attributes, the dataset was randomly
undersampled to result in a balanced subsample made up of 3000
individuals. Secondly, we the German Credit dataset which involves the
task of predicting if bank customers are credit-worthy or not
\protect\hyperlink{ref-germancredit1994}{{[}25{]}}. It consists of 700
positive and 300 negative instances charaterized by 7 numerical and 13
categorical attributes. We process the dataset in two ways: (1) the
values of the ``Personal status and sex'' feature are aggregated by the
two represented genders; (2) the most common values are calculated for
all categorical features such that a feature \(x_d\) with the mode
\(\bar{x}_d\) is transformed into a new binary feature
\(\tilde{x}_d=\mathbb{I}_{x_{d,i}>=\bar{x}_d}\). Binarization ensures
that we can use all counterfactual generators in the benchmark.

\hypertarget{sec-method-metrics}{%
\subsection{Evaluation Metrics}\label{sec-method-metrics}}

The evaluation metric in Equation~\ref{eq-mmd} is computed after every
round \(t=1,...,T\) of the experiment. To assess the statistical
significance of the observed shifts under the null hypothesis that
samples \(X\) and \(\tilde{X}\) were drawn from the same probability
distribution we follow
\protect\hyperlink{ref-arcones1992bootstrap}{{[}26{]}}. To that end, we
combine the two samples and generate a large number of permutations of
\(X + \tilde{X}\). Then, we split the permuted data into two new samples
\(X^\prime\) and \(\tilde{X}^\prime\) having the same size as the
original samples. Then under the null hypothesis we should have that
\(MMD_u^2[F,X^\prime,\tilde{X}^\prime]\) be approximately equal to
\(MMD_u^2[F,X,\tilde{X}]\). The corresponding \(p\)-value can then be
calculated by counting how these two quantities are not equal.

As an alternative to PP MMD we use a pseudo-distance for the
\textbf{Disagreement Coefficient} (Disagreement). This metric was
introduced in \protect\hyperlink{ref-hanneke2007bound}{{[}27{]}} and
estimates \(p(M(x) \neq M^\prime(x))\), that is the probability that two
classifiers do not agree on the predicted outcome for a randomly chosen
sample. Thus, it is not relevant whether the classification is correct
according to the ground truth, but only whether the sample lies on the
same side of the two respective decision boundaries. In our context,
this metric quantifies the overlap between the initial model (trained
before the application of recourse) and the updated model. A
Disagreement Coefficient unequal to zero is indicative of a model shift.
The opposite is not true: even if the Disagreement Coefficient is equal
to zero a model shift may still have occured. This is one reason for why
PP MMD is our our preferred metric.

\hypertarget{sec-empirical}{%
\section{Experiments}\label{sec-empirical}}

@@

\leavevmode\vadjust pre{\hypertarget{ref-gmsc_data}{}}%
\CSLLeftMargin{{[}24{]} }%
\CSLRightInline{K. Competition, {``Give me some credit, improve on the
state of the art in credit scoring by predicting the probability that
somebody will experience financial distress in the next two years.''}
\url{https://www.kaggle.com/c/GiveMeSomeCredit}}

\leavevmode\vadjust pre{\hypertarget{ref-germancredit1994}{}}%
\CSLLeftMargin{{[}25{]} }%
\CSLRightInline{H. Hoffman, {``German credit data,''} 1994.
\url{https://archive.ics.uci.edu/ml/datasets/statlog+(german+credit+data)}}

\leavevmode\vadjust pre{\hypertarget{ref-arcones1992bootstrap}{}}%
\CSLLeftMargin{{[}26{]} }%
\CSLRightInline{M. A. Arcones and E. Gine, {``On the bootstrap of u and
v statistics,''} \emph{The Annals of Statistics}, pp. 655--674, 1992.}

\leavevmode\vadjust pre{\hypertarget{ref-hanneke2007bound}{}}%
\CSLLeftMargin{{[}27{]} }%
\CSLRightInline{S. Hanneke, {``A bound on the label complexity of
agnostic active learning,''} in \emph{Proceedings of the 24th
international conference on machine learning}, 2007, pp. 353--360.}

>>>>

\begin{document}
\maketitle
\begin{abstract}
Existing work on Counterfactual Explanations (CE) and Algorithmic
Recourse (AR) has largely been limited to the static setting: given some
classifier we are interested in finding close, actionable, realistic,
sparse, diverse and ideally causally founded counterfactuals. The
ability of CE to handle dynamics like data and model drift remains a
largely unexplored research challenge at this point. Only one recent
work considers the implications of exogenous domain and model shifts.
This project instead focuses on endogenous dynamics, that is shifts that
occur when AR is actually implemented by a proportion of individuals.
Early findings suggest that the involved shifts may be large with
important implications on the validity of AR and the overall
characteristics of the sample population.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, enhanced, frame hidden, boxrule=0pt, sharp corners]}{\end{tcolorbox}}\fi

@@

\hypertarget{sec-related}{%
\section{Background}\label{sec-related}}

@@

\hypertarget{sec-method}{%
\section{Gradient-Based Recourse Revisited}\label{sec-method}}

In the following we first set out a generalized framework for
gradient-based counterfactual search in Section~\ref{sec-method-general}
to introduce the various counterfactual generators we have chosen to use
in our experiments. We then describe the experimental setup in
Section~\ref{sec-method-2-experiment} and introduce several evaluation
metrics used to benchmark the different generators.

\hypertarget{sec-method-general}{%
\subsection{From individual recourse
\ldots{}}\label{sec-method-general}}

In this work we have chosen to focus on a number of gradient-based
counterfactual generators to investigate the endogenous dynamics we
introduced in Section~\ref{sec-intro}. Gradient-based counterfactual
search is well-suited for differentiable black-box models like deep
neural networks. We can restate Equation~\ref{eq-solution} in a more
general form that encompasses most gradient-based approaches to
counterfactual search:

@@

Here \(\mathbf{s}^\prime=\left\{s_k^\prime\right\}_K\) is the stacked
\(K\)-dimensional array of counterfactual states and
\(f: \mathcal{S} \mapsto \mathcal{X}\) maps from the counterfactual
state space to the feature space.

\hypertarget{towards-collective-recourse}{%
\subsection{\ldots{} towards collective
recourse}\label{towards-collective-recourse}}

@@

\hypertarget{a-note-on-convergence}{%
\subsection{A note on convergence}\label{a-note-on-convergence}}

@@

\hypertarget{sec-method-2}{%
\section{Modeling Endogenous Macrodynamics in Algorithmic
Recourse}\label{sec-method-2}}

In this \ldots{}

\hypertarget{sec-method-2-experiment}{%
\subsection{Simulations}\label{sec-method-2-experiment}}

@@

\hypertarget{sec-method-2-metrics}{%
\subsection{Evaluation Metrics}\label{sec-method-2-metrics}}

The evaluation metric in Equation~\ref{eq-mmd} is computed after every
round \(t=1,...,T\) of the experiment. To assess the statistical
significance of the observed shifts under the null hypothesis that
samples \(X\) and \(\tilde{X}\) were drawn from the same probability
distribution we follow
\protect\hyperlink{ref-arcones1992bootstrap}{{[}24{]}}. To that end, we
combine the two samples and generate a large number of permutations of
\(X + \tilde{X}\). Then, we split the permuted data into two new samples
\(X^\prime\) and \(\tilde{X}^\prime\) having the same size as the
original samples. Then under the null hypothesis we should have that
\(MMD_u^2[F,X^\prime,\tilde{X}^\prime]\) be approximately equal to
\(MMD_u^2[F,X,\tilde{X}]\). The corresponding \(p\)-value can then be
calculated by counting how these two quantities are not equal.

As an alternative to PP MMD we use a pseudo-distance for the
\textbf{Disagreement Coefficient} (Disagreement). This metric was
introduced in \protect\hyperlink{ref-hanneke2007bound}{{[}25{]}} and
estimates \(p(M(x) \neq M^\prime(x))\), that is the probability that two
classifiers do not agree on the predicted outcome for a randomly chosen
sample. Thus, it is not relevant whether the classification is correct
according to the ground truth, but only whether the sample lies on the
same side of the two respective decision boundaries. In our context,
this metric quantifies the overlap between the initial model (trained
before the application of recourse) and the updated model. A
Disagreement Coefficient unequal to zero is indicative of a model shift.
The opposite is not true: even if the Disagreement Coefficient is equal
to zero a model shift may still have occured. This is one reason for why
PP MMD is our our preferred metric.

\hypertarget{sec-empirical}{%
\section{Experiment Setup}\label{sec-empirical}}

\hypertarget{sec-empirical-recourse}{%
\subsection{Recourse Generators}\label{sec-empirical-recourse}}

In the case of the baseline counterfactual generator
\protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}} \(f\) is just
the idendity function and the number of counterfactuals \(K\) is equal
to one. This generator, which we shall refer to as \textbf{Wachter} in
the following, shall serve as the baseline against which all other
gradient-based methodologies will be compared. In particular, we include
include the following generator in our benchmarking exercises: REVISE
\protect\hyperlink{ref-joshi2019towards}{{[}8{]}}, CLUE
\protect\hyperlink{ref-antoran2020getting}{{[}10{]}}, DICE
\protect\hyperlink{ref-mothilal2020explaining}{{[}9{]}} and a greedy
approach that relies on probabilistic models
\protect\hyperlink{ref-schut2021generating}{{[}7{]}}.

\hypertarget{sec-empirical-data}{%
\subsection{Data}\label{sec-empirical-data}}

@@

Additionally, we use two real-world datasets from the Finance domain.
Firstly, we use the Give Me Some Credit dataset which was open-sourced
on Kaggle for the task to predict whether a borrower is likely to
experience financial difficulties in the next two years
\protect\hyperlink{ref-gmsc_data}{{[}26{]}}. Originally consisting of
250,000 instances with 11 numerical attributes, the dataset was randomly
undersampled to result in a balanced subsample made up of 3000
individuals. Secondly, we the German Credit dataset which involves the
task of predicting if bank customers are credit-worthy or not
\protect\hyperlink{ref-germancredit1994}{{[}27{]}}. It consists of 700
positive and 300 negative instances charaterized by 7 numerical and 13
categorical attributes. We process the dataset in two ways: (1) the
values of the ``Personal status and sex'' feature are aggregated by the
two represented genders; (2) the most common values are calculated for
all categorical features such that a feature \(x_d\) with the mode
\(\bar{x}_d\) is transformed into a new binary feature
\(\tilde{x}_d=\mathbb{I}_{x_{d,i}>=\bar{x}_d}\). Binarization ensures
that we can use all counterfactual generators in the benchmark.

\hypertarget{sec-empirical-2}{%
\section{Experiments}\label{sec-empirical-2}}

\hypertarget{endogenous-macrodynamics}{%
\subsection{Endogenous Macrodynamics}\label{endogenous-macrodynamics}}

\hypertarget{potential-mitigation-strategies}{%
\subsection{Potential Mitigation
Strategies}\label{potential-mitigation-strategies}}

@@

\leavevmode\vadjust pre{\hypertarget{ref-arcones1992bootstrap}{}}%
\CSLLeftMargin{{[}24{]} }%
\CSLRightInline{M. A. Arcones and E. Gine, {``On the bootstrap of u and
v statistics,''} \emph{The Annals of Statistics}, pp. 655--674, 1992.}

\leavevmode\vadjust pre{\hypertarget{ref-hanneke2007bound}{}}%
\CSLLeftMargin{{[}25{]} }%
\CSLRightInline{S. Hanneke, {``A bound on the label complexity of
agnostic active learning,''} in \emph{Proceedings of the 24th
international conference on machine learning}, 2007, pp. 353--360.}

\leavevmode\vadjust pre{\hypertarget{ref-gmsc_data}{}}%
\CSLLeftMargin{{[}26{]} }%
\CSLRightInline{K. Competition, {``Give me some credit, improve on the
state of the art in credit scoring by predicting the probability that
somebody will experience financial distress in the next two years.''}
\url{https://www.kaggle.com/c/GiveMeSomeCredit}}

\leavevmode\vadjust pre{\hypertarget{ref-germancredit1994}{}}%
\CSLLeftMargin{{[}27{]} }%
\CSLRightInline{H. Hoffman, {``German credit data,''} 1994.
\url{https://archive.ics.uci.edu/ml/datasets/statlog+(german+credit+data)}}

### 2022-08-11 dev/paper/sections/empirical.qmd
some rejigging of sections

# Experiments {#sec-empirical}

1. Shift of focus from individual to group of individuals (related: https://www.researchgate.net/publication/353073138_Generating_Collective_Counterfactual_Explanations_in_Score-Based_Classification_via_Mathematical_Optimization)
2. Convergence criterium matters: terminating once threshold probability is reached may not be optimal (see e.g. REVISE)
3. Optimizer choice matters: dimensionality is typically low, so no obvious benefit to using ADAM.
    - This might be better placed in JuliaCon proceedings, perhaps backed by small blog post on the matter. 
4. Mitigating strategy: penaliye distance from centroid.

>>>>

# Experiment Setup {#sec-empirical}

## Recourse Generators {#sec-empirical-recourse}

In the case of the baseline counterfactual generator [@wachter2017counterfactual] $f$ is just the idendity function and the number of counterfactuals $K$ is equal to one. This generator, which we shall refer to as **Wachter** in the following, shall serve as the baseline against which all other gradient-based methodologies will be compared. In particular, we include include the following generator in our benchmarking exercises: REVISE [@joshi2019towards], CLUE [@antoran2020getting], DICE [@mothilal2020explaining] and a greedy approach that relies on probabilistic models [@schut2021generating].

Both **REVISE** and **CLUE** search counterfactuals in some latent embedding $S \subset \mathcal{S}$ instead of the feature space directly. The latent embedding is learned by a separate generative model that is tasked with learning the data generating process (DGP) of $X$. In this case $f$ in @eq-general corresponds to the decoder part of the generative model, in other words the deterministic function that maps back from the latent embedding to the feature space. Provided the generative model is well-specified, traversing the latent embedding typically results in realistic and plausible counterfactuals, because they are implicitly generated by the (learned) DGP [@joshi2019towards]. CLUE distinguishes itself from REVISE and other counterfactual generators in that it aims to minimize the predictive uncertainty of the model in question $M$. To quantify predictive uncertainty the authors rely on entropy estimates for probabilistic models. The **Greedy** approach proposed by @schut2021generating also works with the subclass of models $\tilde{\mathcal{M}}\subset\mathcal{M}$ that can produce predictive uncertainty estimates. The authors show that in this setting the complexity penalty $h(\cdot)$ in @eq-general is redundant and meaningful counterfactuals can be generated in a fast and efficient manner through a modified Jacobian-based Saliency Map Attack (JSMA). Finally, **DICE** distinguishes itself from all other generators considered here in that it aims to generate a diverse set of $K>1$ counterfactuals. To this end the authors use a complexity penalty $h(\mathbf{s}^\prime)$ that favours diverse outcomes, in the sense that $s_1, ... , s_K$ look as different from each other as possible.

Our motivation for including these different generators in our analysis, is that they all offer slightly different approaches to generate meaningful counterfactuals for differentiable black-box models. We hypothesize that generating more **meaningful** counterfactuals should mitigate the endogenous dynamics illustrated in @fig-poc in @sec-intro. This intuition stems from the underlying idea that more meaningful counterfactuals are generated by the same or at least a very similar data generating process as the training data. All else equal, counterfactuals that fulfill this basic requirement should be less prone to trigger domain and model shifts.

## Data {#sec-empirical-data}

We have chosen to work with both synthetic and real-world datasets. Using synthetic data allows us to impose distributional properties that may affect the resulting recourse dynamics. Following @upadhyay2021towards we generate synthetic data in $\mathbb{R}^2$ to also allow for a visual interpretation of the results. Real-world data is used in order to assess if endogenous dynamics also occur in higher-dimensional settings.

### Synthetic data

```{julia}
using Plots, PlotThemes
theme(:wong)
catalogue = AlgorithmicRecourseDynamics.Data.load_synthetic()
function plot_data(data,title)
    plt = plot(title=title)
    scatter!(data)
    return plt
end
plts = [plot_data(data,name) for (name, data) in catalogue]
plt = plot(plts..., layout=(1,4), size=(850,200))
savefig(plt, "dev/paper/www/synthetic_data.png")
```

We use four synthetic binary classification datasets consisting of 1000 samples each.^[To see how the data is generated see here: [https://github.com/pat-alt/AlgorithmicRecourseDynamics.jl/blob/main/notebooks/synthetic_datasets.ipynb](https://github.com/pat-alt/AlgorithmicRecourseDynamics.jl/blob/main/notebooks/synthetic_datasets.ipynb)] The datasets are presented in @fig-synthetic-data (see also Appendix A for a formal description). Samples from the negative class are marked in blue while samples of the positive class are marked in orange.

![Synthetic classification datasets used in our experiments.](www/synthetic_data.png){#fig-synthetic-data fig.pos="h" width="8cm" height="2cm"}

Ex-ante we expect to see that Wachter will create a new cluster of counterfactual instances in the proximity of the initial decision boundary. Thus, the choice of a black-box model may have an impact on the paths of the recourse. For generators that use latent space search (@joshi2019towards, @antoran2020getting) or rely on (and have access to) probabilistic models (@antoran2020getting, @schut2021generating) we expect that counterfactuals will end up in regions of the target domain that are densely populated by training samples. Finally, we expect to see the counterfactuals generated by DiCE to be uniformly spread around the feature space inside the target class.

### Real-world data

Additionally, we use two real-world datasets from the Finance domain. Firstly, we use the Give Me Some Credit dataset which was open-sourced on Kaggle for the task to predict whether a borrower is likely to experience financial difficulties in the next two years [@gmsc_data]. Originally consisting of 250,000 instances with 11 numerical attributes, the dataset was randomly undersampled to result in a balanced subsample made up of 3000 individuals. Secondly, we the German Credit dataset which involves the task of predicting if bank customers are credit-worthy or not [@germancredit1994]. It consists of 700 positive and 300 negative instances charaterized by 7 numerical and 13 categorical attributes. We process the dataset in two ways: (1) the values of the "Personal status and sex" feature are aggregated by the two represented genders; (2) the most common values are calculated for all categorical features such that a feature $x_d$ with the mode $\bar{x}_d$ is transformed into a new binary feature $\tilde{x}_d=\mathbb{I}_{x_{d,i}>=\bar{x}_d}$. Binarization ensures that we can use all counterfactual generators in the benchmark.
